---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Lucia Murillo"
date: "2026-05-22"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)
```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col,
  "Penstemon centranthifolius" = pencen_col,
  "Encelia californica" = enccal_col,
  "Salvia leucophylla" = salleu_col,
  "Stipa pulchra" = stipul_col
  
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_minimal())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating new object from drought_exp
drought_exp_clean <- drought_exp |>  
  # making column names lowercase with underscores 
  clean_names() |> 
  # changing the abbreviations to the whole name in new column
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # creating new column with full name for treatment 
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # reordering species levels 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
# reordering water treatment levels 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting for desired columns  
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking-structure

str(drought_exp_clean)
```



Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows
slice_sample(
  drought_exp_clean, # data frame to display 
  n = 10) # displaying 10 rows 
```




# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
#| label: scatterplot

# mase layer ggplot 
ggplot(data = drought_exp_clean,
       # x-xis SLA 
       # x-axis total mass
       mapping = aes(x = sla,
                     y = total_g)) + 
  # scatetrplot 
  geom_point(col = "magenta3")+ 
  # relabelling axes, adding a title 
  labs(x = expression("Specific leaf area (SLA)" ~(cm^3/g)), 
       # could also do "SLA (cm\U00B2 /g)" (UNICODE)
      y = "Total plant mass (g)",
      title = "As SLA increases, total mass also increases")

```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: jitter-treatment

# base layer ggplot 
ggplot(data = drought_exp_clean,
       # x: treatment 
       # y: total mass 
       # color by treatment 
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment
                     )) + 
  # jitter to represent individual plants 
  geom_jitter(height = 0,
              width = 0.1) + 
  # settng custom colors 
  scale_color_manual(
    values = c("Drought stressed" = ds_col,
               "Well watered" = ww_col)
  ) + 
  # relabeling, adding colors 
  labs(x = "Water Treatment",
       y = "Total mass (g)",
       title = "On average, well watered plants have higher mass \nthan drought stressed plants ") +
  # removing legend 
  theme(legend.position = "none") 

```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: jitter-species

# base ggplot 
ggplot(data = drought_exp_clean,
       mapping = aes(x = species_name, # x: species name 
                     y = total_g, # y: total mass 
                     color = species_name # color by species name 
                     )) + 
  geom_jitter(height = 0,
              width = 0.1,
              ) + 
  scale_color_manual(values = plant_colors)+
  # wrapping x-axis text 
  scale_x_discrete(labels = label_wrap(10)) +
  labs(x = "Species",
       y = "Total mass (g)",
       title = "Species differ in total plant mass"
       ) +
  theme(legend.position = "none")
# behind on annotations 

```







# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model-fitting

# model 0: null model
model0 <- lm(
  total_g ~1, # formula: no predictors 
  data = drought_exp_clean # clean data frame
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula: all predictors 
  data = drought_exp_clean
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean)

# model 5: SLA
model5 <- lm(
  total_g ~ sla, 
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species
model7 <- lm(
  total_g ~ species_name, # formula: species only 
  data = drought_exp_clean
)

# ran out of time for annotations alos 
```

## Model diagnostics

```{r}
#| label: check-diagnostucs

# displaying plots in 2x2 grid 
par(mfrow = c(2, 2))
# 
plot(model4)
```

- residuals appear normally distruuted
- residuals appear homoscedastic (have constant variance) 
- no clear outliers in data set influencing model coefficient estimates 



## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating aic

# calculating AIC 
AICc(
  model0,
  model1,
  model2, 
  model3,
  model4,
  model5, 
  model6,
  model7
) |> 
  # arranging by AIC 
  arrange(AICc)

```



Insert code to display the summary for the best model. 

```{r}
#| label: model-summary

summary(model4)

```

- `(intercept)` represents model preductuion for plant mass for Lotus scoparius in the drought stressed treatment 
-> reference levels (dont show up in r output) 
- `water_treatmentWell watered`: model predicyion for differnce in Lotus scoparious mass between well watered and drought stressed treatment. 
- `species_namePenstemon centrathifolius`: model pred for the diff between Penstemon mass in the well watered treatment and lotus mass in teh well watered treatment. 
- `species_nameEschscholzia californica`: model prediction for the diff between E californica mass in the well watered treatment and lotus mass in well watered 



## Stop and think: what does this model mean?  

**What is the best model?**

The best model inclydes water treatment and species as predictors 

**How much variation in the response (total mass, in grams) does this model explain?**

72% of the variation in plant mass is explained bbya model with both of these predictors. 

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- treatment influences plant mass. 
- species influences plant mass. 

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions

preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x, 
         species_name = group)


```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final-figure

ggplot(data = drought_exp_clean,
       mapping = aes(x = water_treatment, 
                     y = total_g,
                     color = water_treatment)) + 
  geom_jitter(height = 0, 
              width = 0.1,
              alpha = 0.8) +
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment,
                               y = predicted,
                               ymin = conf.low,
                               ymax = conf.high)) + 
  facet_wrap(~species_name) + 
  theme(legend.position = "none")


```


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







